Spatial Analysis

### 2021-06-08 Scripts/spatial_analisis.Rmd
End to end analysis done  for black seabass

---
title: "Spatial Analysis"
# output: html_notebook
output: html_document
---

```{r setup, include=FALSE}
MyFunctions::my_lib(c("ggmap","sf","tidyverse","tools","readr","data.table","maps","viridis"))
library(MyFunctions)
```

@@

## Ocean adapt data

- Using only "Northeast US Fall and Spring surveys for now

ocean_data <- readRDS("/Volumes/Enterprise/Data/pinskylab-OceanAdapt-966adf0/data_clean/dat_exploded.rds") %>% 
  filter(spp == "Centropristis striata",
       region %in% c("Northeast US Fall" , "Northeast US Spring")) #No more seasons

@@

@@

States <- c("maine", "new hampshire", "massachusets", "connecticut", "rhode island", "new york", "new jersey", "delaware", "maryland", "virginia", "north carolina")

@@

@@

## Create basis sf

In this first step we first combine the North East EEZ shapefile (SAU) with the state waters sf provided by [NOAA](link) and girded the area following these steps:

@@

- 2. Grid that buffer on a 0.3 ressolution

@@

### 1. Expand Spatial regions (a.k.a  buffers)

A buffer of *410000* m (410 KM / 221) overshoots the EEZ a bit, but is eventually cropped

@@

### 2. Expand grid within buffer

Here we expand a grid within the buffer so we can estimate the catch within

@@

#### 2.1 Merge grid and buffers

Here we converted the grid to `sf`, merged with the buffered states and filter out everything outside the states polygon

@@

gridExtra::grid.arrange(
  # Overall (overlapping) position
  ggplot(grid_sf) +
    geom_sf(aes(color = state)) +
    geom_sf(data = eez_sf, aes(),fill =NA) + 
    theme(legend.position = ""),
  # Showing each state separatley
  ggplot() +
    geom_sf(data = grid_sf, aes(color = state),size = 0.1) +
    facet_wrap(~state) +
    theme(legend.position = "top"),
  nrow = 1)

@@

### 3. Crop buffers to EEZ

```{r buff_to_eez, eval = T, echo = T, message = F, warning = F}

@@

# write_sf(grid_eez_sf, "../Results/Partial/grid_eez_sf.shp")

# Get final df with index
grid_eez_df <- as.data.frame(grid_eez_sf) %>%
  select(state,index)

# write_csv(grid_eez_df, "../Results/Partial/grid_eez_df.csv")

# Plot to make sure makes sense (Picasso syle)
# grid_eez_sf %>%
#   st_simplify(preserveTopology = TRUE, dTolerance = 10000) %>% #0.1 for paper
#   ggplot() +
#   geom_sf(aes(color = state))

@@

## Map of smapling points

In this case the color gradient represents wtcpue (wet tones catch per unit of effort, haul hours) from 1972 to 2019. Grey dots represent `wtcpue = 0`

```{r survey_map, eval = F, echo = T, fig.width = 12, fig.height = 8}

ggplot(land_sf) +
  geom_sf() +
  geom_sf(data = eez_sf, aes(), fill = "white") +
  geom_point(data = subset(ocean_data, wtcpue = 0),
             aes(
               x = lon,
               y = lat
             ),
             color = "grey95",
             size = 1
  ) +
  geom_point(data = subset(ocean_data, wtcpue > 0),
             aes(
               x = lon,
               y = lat,
               color = log10(wtcpue)
             ),
             size = 1
  ) +
  scale_color_distiller(palette = "Spectral", 
                       guide_legend(title = "WCPUE per Haul (log10)")) + 
  coord_sf(xlim = c(-76, -65),ylim = c(35, 45)) +
  MyFunctions::my_ggtheme_m() +
  facet_wrap(~region)

# Interpolation

Here we [interpolate](https://swilke-geoscience.net/post/spatial_interpolation/) the survey data within the grid. We

## Prepare data

@@

```{r prepare_data, eval = T, echo = T}

# Subset data
subset_data <- ocean_data %>% 
  filter(
    region == "Northeast US Spring",
    wtcpue > 0)

# Get grid
grd_template <- data.frame(index = grid_sf$index,
                       state = grid_sf$state,
                       st_coordinates(grid_sf))

## Fun. interpol

@@

tis <- function(input_data,yr){

@@

  data <- input_data %>% 
    filter(year == yr)

  if(nrow(data) == 0){
    return(tibble())
  }else{
    # Triangular Irregular Surface
    fit_tin <- interp::interp( # using {interp}
      x = data$lon,           # the function actually accepts coordinate vectors
      y = data$lat,
      z = data$wtcpue,
      xo = grd_template$X,     # here we already define the target grid
      yo = grd_template$Y,
      output = "points"
    ) %>% 
      bind_cols() %>% 
      bind_cols(grd_template) %>%
      mutate(year = yr) %>% 
      select(index, state, year, lon=x, lat=y, value = z)

    return(fit_tin)
  }
}

@@

## Control pannel

@@

@@

historic_tif <- bind_rows(
  lapply(years,tis,input_data = subset_data)
)

write_csv(historic_tif,
          paste0(my_path("R",),"interpolated_data.csv"))

@@

@@

@@

```{r area_plot, eval = T, echo = T}

 data_grid <- historic_tif %>% 
  # left_join(periods) %>% 
  # group_by(state,lat,lon,period) %>% 
  group_by(state,lat,lon) %>% 
  summarise(mean = mean(value,na.rm= T), .groups = "drop") %>% 
  rename(ID = state)

ggplot(data_grid) +
  geom_tile(
    aes(
      x = lon,
      y = lat,
      fill = mean,
      col = mean
    )
  ) +
  facet_wrap(~ ID) +
  scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  scale_color_viridis("Mean Proportion", alpha = 0.8)

@@

```{r area_plot, eval = T, echo = T}

total_fited <- historic_tif %>% 
  group_by(year) %>% 
  summarise(total_value = sum(value,na.rm=T))

state_fit <- historic_tif %>% 
  group_by(state,year) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = "year") %>%
  mutate(percentage = state_value/total_value*100) %>% 
  left_join(periods) %>% 
  group_by(ID = state,period) %>% 
  summarise(mean_per = round(mean(percentage)))

@@

land_sf %>% 
  left_join(state_fit,
            by = "ID") %>% 
  ggplot() +
  geom_sf(aes(fill = mean_per)) +
  viridis::scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  facet_wrap(~period, nrow = 1) +
  my_ggtheme_m()

@@

```{r area_plot, eval = T, echo = T}

total_fited <- historic_tif %>% 
  group_by(year) %>% 
  summarise(total_value = sum(value,na.rm=T))

# group by state
state_fit <- historic_tif %>% 
  group_by(state,year) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = "year") %>%
  mutate(percentage = state_value/total_value*100)

p <- ggplot(state_fit) +
  geom_area(
    aes(
      x = year,
      y = round(percentage),
      fill = state
    )
  ) +
  viridis::scale_fill_viridis(discrete = T, alpha = 0.8) +
  MyFunctions::my_ggtheme_p()

plotly::ggplotly(p)

>>>>

---
title: "Spatial Analysis"
output: 
 # html_document
  html_notebook:
  fig_width: 6
  fig_height: 4
  toc: false
  toc_depth: 3
  code_folding: hide
  toc_float:
    collapsed: false
---

```{r setup, include=FALSE}
library(MyFunctions)
MyFunctions::my_lib(c("ggmap","sf","tidyverse","tools","readr","data.table","maps","viridis","wesanderson","knitr","kableExtra","plotly"))
```

@@

@@

We use three different shapefiles for the continental U.S. land mass, the State waters of maine, new hampshire, massachusets, connecticut, rhode island, new york, new jersey, delaware, maryland, virginia, north carolina and the North East EEZ.

1.- **Land shapefile;** covers the US land territory for visualization. Data provided from the `map` package.

2.- **State waters;** covers the state waters of the NE US states. Data from [data.gov](https://catalog.data.gov/dataset/federal-and-state-waters).

*License: No license information was provided. If this work was prepared by an officer or employee of the United States government as part of that person's official duties it is considered a U.S. Government Work.*

3.- **EEZ shapefile;** Used the *Sea Around Us* shapefle updated to June 2016.

@@

States <- c("maine", "new hampshire", "massachusetts", "connecticut", "rhode island", "new york", "new jersey", "delaware", "maryland", "virginia", "north carolina","pennsylvania")

@@

unique(land_sf$ID)

@@

## Spatial component

### Create base shapefile for computation

As a first step we need to divide the NE US EEZ among the different states. For that we expanded state waters up to the 200 nautical miles to then estimate the percentage that each expanded-state-waters occupied. Note that in all cases these areas overlapped and percentages accounted for it. We did this by following these steps:

@@

- 2. Grid that buffer on a 0.3 resolution

@@

#### 1. Expand Spatial regions (a.k.a  buffers)

We set a buffer of *410000* m (410 km, ~ 221 nm) that overshoots the EEZ a bit, but is eventually cropped

@@

#### 2. Expand grid within buffer

Here we expand a grid within the buffer so we can estimate the proportion of each state

@@

##### 2.1 Merge grid and buffers

Once we have a gridded area, we converted the grid to a `sf` so we can merge it with the buffered states and finally filter out everything outside the states polygon

@@

gridExtra::grid.arrange(
  # Overall (overlapping) position
  ggplot(grid_sf) +
    geom_sf(aes(color = state), alpha = 0.3) +
    geom_sf(data = eez_sf, aes(),fill =NA) + 
    theme(legend.position = ""),
  # Showing each state separatley
  ggplot() +
    geom_sf(data = grid_sf, aes(color = state),size = 0.1, alpha = 0.3) +
    facet_wrap(~state) +
    theme(legend.position = "top"),
  nrow = 1)

@@

#### 3. Crop buffers to EEZ

Finally, we crop the grided buffers to within the EEZ to capture the actual water space.

*Note:* This step takes quite a while because of the size of the EEZ shapefile. No, you cannot use `st_simplify()` here

```{r buff_to_eez, eval =T, echo = T, message = F, warning = F}

@@

# write_sf(grid_eez_sf, paste0(my_path("D", "Spatial"),"grid_eez_sf.shp"))

# Get final df with index
grid_eez_df <- as.data.frame(grid_eez_sf) %>%
  select(state,index) %>% 
  left_join(ne_grid, 
            by = "index")

# write_csv(grid_eez_df, paste0(my_path("R", "Partial"),"grid_eez_df.csv"))

# Plot to make sure makes sense (Picasso syle)
grid_eez_sf %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 10000) %>% #0.1 for paper
  ggplot() +
  geom_sf(aes(color = state), alpha = 0.3)

@@

## Interpolation rutine

Here we [interpolate](https://swilke-geoscience.net/post/spatial_interpolation/) the survey data within the previously created grid.

@@

### Function for interpolation

@@

tis <- function(input_data, grid, yr, taxa, reg){

@@

  data <- input_data %>% 
    filter(year == yr,
           spp == taxa,
           region == reg
    ) %>% 
    filter(wtcpue > 0)

    if(nrow(data) == 0){
      return(tibble())
    }else{

      # Triangular Irregular Surface
      fit_tin <- interp::interp( # using {interp}
        x = data$lon,           # the function actually accepts coordinate vectors
        y = data$lat,
        z = data$wtcpue,
        xo = grid$lon,     # here we already define the target grid
        yo = grid$lat,
        output = "points"
      ) %>% 
        bind_cols() %>% 
        bind_cols(grid) %>%
        mutate(year = yr,
               region = reg,
               spp = taxa) %>% 
        select(index, state, year, region, spp, lon=x, lat=y, value = z)

    }

    return(fit_tin)
  }

@@

### Ocean adapt data

- Using only the Northeast US Fall and Spring bottom trawl survey data for now

ocean_data <- readRDS("/Volumes/Enterprise/Data/pinskylab-OceanAdapt-966adf0/data_clean/dat_exploded.rds") #%>% 
  # filter(spp == "Centropristis striata",
       # region %in% c("Northeast US Fall" , "Northeast US Spring")) #No more seasons

subset_data <- ocean_data %>% 
  filter(spp == "Centropristis striata",
       region %in% c("Northeast US Fall" , "Northeast US Spring")
       )

ggplot() +
  geom_point(data = subset(subset_data, wtcpue = 0.0),
             aes(
               x = lon,
               y = lat
             ),
             color = "grey95"
  ) +
  geom_point(data = subset(subset_data, wtcpue > 0),
             aes(
               x = lon,
               y = lat,
               color = log10(wtcpue)
             ),
             size = 1
  ) +
  scale_color_distiller(palette = "Spectral", 
                        guide_legend(title = "WCPUE per Haul (log10)")) + 
  coord_sf(xlim = c(-76, -65),ylim = c(35, 45)) +
  MyFunctions::my_ggtheme_m() +
  facet_wrap(~region)

### Control Pannel

Load required data. Note that some of it has been previously created

@@

# Load grid df
grid_eez_df <- my_path("D","Spatial","grid_eez_df.csv", read = T)

@@

# taxa
spp <- "Centropristis striata"

# regions
regions <- c("Northeast US Fall" , "Northeast US Spring")

### Run routine

```{r run_routine, eval = T, echo = T}

# Run for a single region
# historic_tif <- bind_rows(
  # lapply(years,tis,input_data = ocean_data, grid = grid, taxa = "Centropristis striata", reg = regions[2])
# )

# Simple for-loop for running both regions (seasons)
for(r in 1:2){

  partial_df <- bind_rows(
    lapply(years,tis,input_data = ocean_data, grid = grid, taxa = spp, reg = regions[r])
  )

  if(r == 1){
    historic_tif <- partial_df
  }else{
    historic_tif <- bind_rows(historic_tif,partial_df)
  }

}

# Save data for future plotting
# write_csv(historic_tif,
          # paste0(my_path("R","Partial"),"interpolated_data.csv"))

# Show data in notebook
head(historic_tif,10) %>% 
   kable(
    format = "html",
    caption = "Extrapolated data") %>% 
  kableExtra::kable_minimal() %>% 
  kableExtra::kable_styling() %>%
  kableExtra:: scroll_box(width = "100%", height = "300px")

@@

# Load interpolated data
historic_tif <- my_path("R","Partial","interpolated_data.csv",read = T)

@@

# State pallet

state_pallet <- c(wes_palette(n = 5, name = "Darjeeling1"),
                               wes_palette(n = 5, name = "Darjeeling2"),
                               wes_palette(n = 3, name = "Royal1")
                  )

@@

```{r area_map_state, eval = T, echo = T, fig.width=10, fig.height=12}

 data_grid <- historic_tif %>% 
  # left_join(periods) %>% 
  # group_by(state,lat,lon,period) %>% 
  group_by(state,lat,lon,region) %>% 
  summarise(mean = mean(value,na.rm= T), .groups = "drop") %>% 
  rename(ID = state)

ggplot(data_grid) +
  geom_tile(
    aes(
      x = lon,
      y = lat,
      fill = mean,
      col = mean
    )
  ) +
  facet_wrap(~ ID  +  region,
             ncol = 4) +
  scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  scale_color_viridis("Mean Proportion", alpha = 0.8)

@@

```{r area_map, eval = T, echo = T, fig.width = 10}

total_fited <- historic_tif %>% 
  group_by(year,region) %>% 
  summarise(total_value = sum(value,na.rm=T))

state_fit <- historic_tif %>% 
  group_by(state,year,region) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = c("year","region")) %>%
  mutate(percentage = state_value/total_value*100) %>% 
  left_join(periods) %>% 
  group_by(ID = state,period,region) %>% 
  summarise(mean_per = round(mean(percentage)))

@@

land_sf %>% 
  left_join(state_fit,
            by = "ID") %>% 
  ggplot() +
  geom_sf(aes(fill = mean_per)) +
  viridis::scale_fill_viridis("Mean Proportion", alpha = 0.8) +
  facet_wrap(~region + period, nrow = 2) +
  my_ggtheme_m()

@@

```{r area_plot, eval = T, echo = T, fig.height = 10, fig.width = 10}

total_fited <- historic_tif %>% 
  group_by(year,region) %>% 
  summarise(total_value = sum(value,na.rm=T))

# group by state
state_fit <- historic_tif %>% 
  group_by(state,year,region) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = c("year","region")) %>%
  mutate(percentage = state_value/total_value*100) %>% 
  group_by(state,region)

p <- ggplot(state_fit) +
  geom_area(
    aes(
      x = year,
      y = round(percentage),
      fill = state
    )
  ) +
  ylab("Percentage (%)") +
  # viridis::scale_fill_viridis(discrete = T, alpha = 0.8) +
  scale_fill_manual(values = state_pallet) +
  MyFunctions::my_ggtheme_p() +
  facet_wrap(~region, ncol = 1)

ggplotly(p,
         dynamicTicks = TRUE) %>% 
  layout(hovermode = "x") %>% 
  # add_trace() %>% 
  rangeslider()

```
## Area plot (running)

This graph shows the proportion of each State smoothed over a 10 years running mean. It helps seeing the trends better (I Think...)

```{r area_lm_plot, eval = T, echo = T, fig.height = 10, fig.width = 10}

# group by state
state_fit <- historic_tif %>% 
  group_by(state,year,region) %>% 
  summarise(state_value = sum(value,na.rm= T), .groups = "drop") %>% 
  left_join(total_fited,
            by = c("year","region")) %>%
  mutate(percentage = state_value/total_value*100) %>% 
  group_by(state,region) %>% 
  mutate(RMean = zoo::rollmean(x = percentage, 
                            10,
                            fill = NA,
                            na.rm=T)
    )

# Plot me
p <- ggplot(state_fit) +
  geom_area(
    aes(
      x = year,
      y = round(RMean),
      fill = state
    )
  ) +
  ylab("10 yrs running average (%)") +
  # viridis::scale_fill_viridis(discrete = T, alpha = 0.8) +
  scale_fill_manual(values = state_pallet) +
  MyFunctions::my_ggtheme_p() +
  facet_wrap(~region, ncol = 1)

suppressWarnings(
ggplotly(p,
         dynamicTicks = TRUE) %>% 
  layout(hovermode = "x") %>% 
  # add_trace() %>% 
  rangeslider()
)
